PanLex

### 2013-01-01 paper/2012/SWJ-PanLex/paper.tex
Reworking the vocabulary design section

\newcommand{\claus}[1]{\todo[inline]{[Claus: #1]}}
%\newcommand{\textrdf}[1]{\texttt{#1}}
\newcommand{\textrdf}[1]{\emph{#1}}

\usepackage{listings}
%\usepackage{verbatim}
\lstset{%
    numberbychapter=false,
    numbers=left,
    numberstyle=\tiny,
    basicstyle=\footnotesize\ttfamily,
    tabsize=2,
    framexleftmargin=2pt,
    captionpos=b,
    frame=single,
    breaklines=true
}
\lstdefinestyle{rdf}{numberblanklines=true, morekeywords={}}
\lstdefinestyle{sparql}{basicstyle=\ttfamily\scriptsize,numberblanklines=true,
    morekeywords={SELECT,OPTIONAL,FROM,DISTINCT,a,WHERE,FILTER,GROUP,ORDER,LIMIT,BY,IN,AS},
    emph={r,pub,aairObject,verb,person,bday,s,p,o},emphstyle=\textit
}
\lstdefinestyle{turtle}{basicstyle=\ttfamily\tiny ,numberblanklines=true,
    morekeywords={a, @prefix},
    morecomment=[s][\textrm]{<}{>},
    morecomment=[s][\textit]{"}{"},
}

@@

\section{Introduction}
% \begin{itemize}
% \item introduce PanLex project
% \item @author: Jonathan
% \item Motivation: About approximately 6900 active languages (\todo{source seems to be http://www.amazon.com/Ethnologue-Languages-M-Paul-Lewis/dp/1556712162}), Existing translation only between relatively few language pairs (dictionaries).
% \item Transitive word-by-word translations 
% * high density and low density languages
% * decline in linguistic diversity
% * These combined forces have led to predictions that something between half and 90%
%   of the world's living languages will die within the next century (Woodbury 2006; UNESCO 2003: 2).
%   Weaker forms of these forces appear to be shrinking the use of medium-density languages in science,
%   diplomacy, business, and other domains (Phillipson 2008).
% * When languages die, they argue, the world loses
%   (1) irreplaceable knowledge of history, medicine, nature, and productive methods encoded in languages'
% * extraction + integration
% \end{itemize}
At present, there are about 7,000 living languages in the world\footnote{See \url{http://www.sil.org/iso639-3/download.asp} for a list of registered languages}.
Nonetheless, some experts claim that processes such as nation-state consolidation and globalization are producing language attrition so rapidly that up to 90\% of all languages alive today will be extinct within a century~\cite{endang_lang}.
Theorists of biolinguistic diversity~\cite{nettle} argue that the loss of
language diversity, the loss of human biological knowledge, and the loss of
species diversity are mutually supportive and thus that language preservation
and revitalization are essential to the preservation of biological diversity.
The vision of the PanLex project is to save all these thousands of languages,
especially those low-density ones that are threatened by extinction,
by supporting their use in global communication. This requires panlingual
translation: translation from any language of the world into any other.
One of the crucial components of panlingual translation of discourses is
panlingual lexical translation. PanLex is designed to support that component.
It documents the known lexical translations (translations of lexemes) among all
languages.

@@

\begin{itemize}
  \item \emph{Expressions} are lemmas. For example, \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
  Expressions are always given in a language variety and can only be given once per language variety.
  They were extracted from the above mentioned mono- or bilingual dictionaries or from other sources.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language Varieties} within a language are further distinguished by arbitrary unique integer identifiers.
  For example, ``eng'' is the ISO 639-3 code for the English language.
  Panlex defines various varieties of it, including ``English'' (eng-000), ``Simple English'' (eng-001) and ``British English'' (eng-005).
  Their labels, when possible, are autonyms, written in the native writing system.
  So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription. %  and one can use variety names like ``{زبان اوستایی}''.
  Additionally these variety identifiers and labels do not have to be registered like the IETF BCP 47 language subtags\footnote{\url{http://www.iana.org/assignments/language-subtag-registry}}.
  \item \emph{Meanings} of expressions are distinguished by editors based on their
  interpretation of an information source.
  Information sources can be dictionaries, thesauri, etc., and also personal knowledge. 
  The pair comprised of the editor and the information source is referred to as a meaning's \emph{approver}.
  Editors may add informal definitions to meanings in any language variety.
  \item \emph{Definitions} are optional descriptions of a meaning.
  They are given in a certain language variety.
  A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}, marked as a definition in the ``English'' language variety.
  \item \emph{Denotations} relate expressions to \emph{meanings}.
  So homonyms are represented by one expression being connected to different meanings.
  Denotations may be enriched with part-of-speech tags selected from a closed list based on the
  Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
  For example, \emph{fall} can be a verb or a noun for autumn.
  Furthermore, it is possible to generically tag denotations with metadata
  using key-value pairs.
  For instance, an English expression \emph{pig} when synonymous with
  \emph{police officer} could be annotated with \emph{pragmatics=vulgar}.
  \item \emph{Users} have editorial privileges over the language varieties and the approvers that they define.
  \item \emph{Licenses} are also considered by the PanLex project.
  At present there are ten different license categories an approver can be annotated with.
  They are \emph{public domain}, \emph{Creative Commons}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License}, \emph{GNU Lesser General Public License}, \emph{GNU Free Documentation License}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
\end{itemize}

@@

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
The schema described above was adopted to create RDF triple structures.
Since the relational schema is a good starting point most of the relations were taken over introducing a PanLex RDF vocabulary.
Therefore the namespace \texttt{<http://panlex.org/plx/>} is used, abbreviated by the \texttt{plx} prefix.

The structure of \emph{approvers} is nearly the same as in the database schema.
All approvers share the \texttt{plx:Approver} class.
The characteristics of an approver are described using mainly triples with literal objects.
These are for example \texttt{plx:registrationDate} pointing to an \texttt{xsd:date} literal, \texttt{dc:title} to assign the title of a source, \texttt{dc:creator} to give an \texttt{xsd:string} containing the author's name etc.
All properties defined for \texttt{plx:Approver} instances are shown in \autoref{tbl:vocabulary}.

Besides the approvers there are also \emph{languages} and \emph{language varieties} taken over into the RDF vocabulary.
They are of the classes \texttt{plx:Language} and \texttt{plx:LanguageVariety} respectively.
Language varieties have a name given by the \texttt{rdfs:label} property and belong to a certain language referred to via \texttt{plx:languageVarietyOf}.
For those languages that have an ISO 639-3 language code, this code is referenced via the \texttt{plx:iso639-3Code} property.
The equivalent ISO 639-1 code can be retrieved using the \texttt{plx:iso639-1Code} property.

\emph{ISO 639-1} and \emph{ISO 639-3 codes} are of the classes \texttt{plx:Iso639-1Code} and \texttt{plx:Iso639-3Code} respectively.
These entities are mainly used for interlinking to the Lexvo dataset\footnote{\url{http://lexvo.org/ontology}}.
See \autoref{sec:linking} for more details.

As in the original PanLex database there are also \emph{expressions} used in the RDF dataset.
These resources have an \texttt{rdfs:label} containing the textual representation of the considered expression.
Additionally there is a normalized version of the label, reachable via the \texttt{plx:degradedText} property.
Expression entities also have an assigned language variety (\texttt{plx:languageVariety}).
It refers to the language variety the textual representation is written in.

@@

Since the PanLex project compiled its database by extracting data from different sources, the licenses of these sources were considered, too.
At present, we support the different license categories given in the database by creating resources of the \texttt{plx:License} class.
In the RDF dataset these have an additional \texttt{rdfs:label} property giving the names or descriptions of the licenses or license categories.
\begin{table}
\begin{tabular}{ll}
Class & Properties \\
\toprule
\texttt{plx:Approver} & \texttt{plx:registrationDate} \\
                      & \texttt{rdfs:label} \\
                      & \texttt{foaf:homepage} \\
                      & \texttt{plx:isbn} \\
                      & \texttt{dc:creator} \\
                      & \texttt{dc:title} \\
                      & \texttt{dc:publisher} \\
                      & \texttt{plx:yearOfPublication} \\
                      & \texttt{plx:quality} \\
                      & \texttt{plx:license} \\
\midrule
\texttt{plx:Language}
                      & \texttt{plx:iso639-3Code} \\
                      & \texttt{plx:iso639-1Code} \\
\midrule
\texttt{plx:LanguageVariety}
                      & \texttt{plx:languageVarietyOf} \\
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:Iso639-1Code} & \\
\midrule
\texttt{plx:Iso639-3Code} & \\
\midrule
\texttt{plx:Expression}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{rdfs:label} \\
                      & \texttt{plx:degradedText} \\
\midrule
\texttt{plx:Meaning}
                      & \texttt{plx:approver} \\
                      & \texttt{plx:meaningDefinition} \\
                      & \texttt{plx:identifier} \\
\midrule
\texttt{plx:Definition}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:Denotation}
                      & \texttt{plx:denotationMeaning} \\
                      & \texttt{plx:denotationExpression} \\
                      & \texttt{plx:wordClass} \\
\midrule
\texttt{plx:WordClass}  
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:License}  & \texttt{rdfs:label} \\
\bottomrule
\end{tabular}
\caption{Classes and properties used in the PanLex RDF vocabulary}
\label{tbl:vocabulary}
\end{table}

An overview of the PanLex vocabulary is shown in \autoref{fig:vocabulary}.%
\begin{figure}
\includegraphics[width=0.47\textwidth]{images/pdf/ontology.pdf}
\caption{Overview of the PanLex RDF vocabulary}
\label{fig:vocabulary}
\end{figure}
There, any \texttt{rdf:type} triples as well as facts assigning ISO 639-1 and ISO 639-3 labels are omitted for clarity.

\subsection{Transformation workflow}
\label{sec:conversion}
\todo{how often in PanLex updated?}
The size of the PanLex database (including indexes) is approximately 18 GB,
which is big enough to make recurrent RDF conversion cumbersome:
A full RDF conversion bloats the dataset and results in data duplication.   
Using conventional hardware, it takes impractically long
to convert all the data. This makes
testing and debugging and fixing issues in the modelling or data conversion of the
data very time-consuming.
As the PanLex data already resides in a relational database, the use of a virtual
RDB 2 RDF mapping solution is a natural choice. The \emph{Sparqlify Mapping Language}
offers a very easy-to-use mapping language.
Essentially, these mappings consist of three clauses:
\begin{itemize}
  \item \emph{From} Specifies the logical SQL table (i.e. table, view or query) underlying the Sparqlify-ML view.
  \item \emph{With} Defines SPARQL variables by means of expressions over relational columns that yield RDF terms.
  \item \emph{Construct} The template (i.e. the set of triple patterns) to be generated in this view, based on the SPARQL variable definitions.
\end{itemize}
~\autoref{fig:ex:sparqlify-ml} shows an example of a Sparqlify-ML view for the
languages in PanLex: From each row of the \emph{i1} table three resources are
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language}in the PanLex and the \emph{schema.org} namespace.
\todo{Maybe we should link to lexvo directly.}
This view-based approach also demonstrates that changing the vocubulary or
adding support for new ones does not require an ETL process, and can therefore be done with little effort.
%PanLex's languages  iso 639-3 code 
%typed with plx:Language
%resources are created for
%the languages in PanLex, linked with \emph{lexvo}, and

@@

\begin{figure}
\centering
\begin{lstlisting}
Create View i1 As
  Construct {
    ?lang a plx:Language ;
          a <http://schema.org/Language> ;
          plx:iso639-3Code ?iso3 .

>>>>

\newcommand{\claus}[1]{\todo[inline]{[Claus: #1]}}
%\newcommand{\textrdf}[1]{\texttt{#1}}
\newcommand{\vocab}[1]{\emph{#1}}

\usepackage[scaled]{beramono}
\newcommand\Small{\fontsize{9}{9.2}\selectfont}
%\newcommand*\LSTfont{\Small\ttfamily\SetTracking{encoding=*}{-60}\lsstyle}

\usepackage{listings}
%\lstset{language=Python}
%\lstset{%
% morekeywords={Create,View,As,Construct,With,From}
%}%
\lstset{emph={%  
    Create,View,As,Construct,With,From%
    },emphstyle={\color{blue}\bfseries}%
}%

%\usepackage{verbatim}
\lstset{%
    numberbychapter=false,
    numbers=left,
    numberstyle=\tiny,
    basicstyle=\footnotesize\ttfamily,
    tabsize=2,
    framexleftmargin=2pt,
    captionpos=b,
    frame=single,
    breaklines=true
}
\lstdefinestyle{rdf}{numberblanklines=true, morekeywords={}}
\lstdefinestyle{sparql}{basicstyle=\ttfamily\scriptsize,numberblanklines=true,
    morekeywords={SELECT,OPTIONAL,FROM,DISTINCT,a,WHERE,FILTER,GROUP,ORDER,LIMIT,BY,IN,AS},
    emph={r,pub,aairObject,verb,person,bday,s,p,o},emphstyle=\textit
}
\lstdefinestyle{turtle}{basicstyle=\ttfamily\tiny ,numberblanklines=true,
    morekeywords={a, @prefix},
    morecomment=[s][\textrm]{<}{>},
    morecomment=[s][\textit]{"}{"},
}

@@

\section{Introduction}
% \begin{itemize}
% \item introduce PanLex project
% \item @author: Jonathan
% \item Motivation: About approximately 6900 active languages (\todo{source seems to be http://www.amazon.com/Ethnologue-Languages-M-Paul-Lewis/dp/1556712162}), Existing translation only between relatively few language pairs (dictionaries).
% \item Transitive word-by-word translations 
% * high density and low density languages
% * decline in linguistic diversity
% * These combined forces have led to predictions that something between half and 90%
%   of the world's living languages will die within the next century (Woodbury 2006; UNESCO 2003: 2).
%   Weaker forms of these forces appear to be shrinking the use of medium-density languages in science,
%   diplomacy, business, and other domains (Phillipson 2008).
% * When languages die, they argue, the world loses
%   (1) irreplaceable knowledge of history, medicine, nature, and productive methods encoded in languages'
% * extraction + integration
% \end{itemize}
At present, there are about 7,000 living languages in the world\footnote{See \url{http://www.sil.org/iso639-3/download.asp} for a list of registered languages}.
Nonetheless, some experts claim that processes such as nation-state consolidation and globalization are producing language attrition so rapidly that up to 90\% of all languages alive today will be extinct within a century~\cite{endang_lang}.
Theorists of biolinguistic diversity argue that the loss of
language diversity, the loss of human biological knowledge, and the loss of
species diversity are mutually supportive and thus that language preservation
and revitalization are essential to the preservation of biological diversity~\cite{nettle}.
The vision of the PanLex project is to save all these thousands of languages,
especially those low-density ones that are threatened by extinction,
by supporting their use in global communication. This requires panlingual
translation: translation from any language of the world into any other.
One of the crucial components of panlingual translation of discourses is
panlingual lexical translation. PanLex is designed to support that component.
It documents the known lexical translations (translations of lexemes) among all
languages.

@@

\begin{itemize}
  \item \emph{Expressions} are lemmas. For example, \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
  Expressions are always given in a language variety and can only be given once per language variety.
  They were extracted from the afore mentioned mono- or bilingual dictionaries or from other sources.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language Varieties} within a language are further distinguished by arbitrary unique integer identifiers.
  For example, ``eng'' is the ISO 639-3 code for the English language.
  Panlex defines various varieties of it, including ``English'' (eng-000), ``Simple English'' (eng-001) and ``British English'' (eng-005).
  Their labels, when possible, are autonyms, written in the native writing system.
  So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription. %  and one can use variety names like ``{زبان اوستایی}''.
  Additionally these variety identifiers and labels do not have to be registered like the IETF BCP 47 language subtags\footnote{\url{http://www.iana.org/assignments/language-subtag-registry}}.
  \item \emph{Meanings} of expressions are distinguished by editors based on their
  interpretation of an information source.
%  Information sources can be dictionaries, thesauri, etc., and also personal knowledge. 
  The pair comprised of the editor and the information source is referred to as a meaning's \emph{approver}.
  Editors may add informal definitions to meanings in any language variety.
  \item \emph{Definitions} are optional descriptions of a meaning.
  They are given in a certain language variety.
  A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}, marked as a definition in the ``English'' language variety.
  \item \emph{Denotations} relate expressions to meanings.
  So homonyms are represented by one expression being connected to different meanings.
  Denotations may be enriched with part-of-speech tags selected from a closed list based on the
  Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
  For example, \emph{fall} can be a verb or a noun for autumn.
  Furthermore, it is possible to generically tag denotations with metadata
  using key-value pairs.
  For instance, an English expression \emph{pig} when synonymous with
  \emph{police officer} could be annotated with \emph{pragmatics=vulgar}.
  \item \emph{Users} have editorial privileges over the language varieties and the approvers that they define.
  \item \emph{Licenses} are also considered by the PanLex project.
  At present there are ten different license categories an approver can be annotated with.
  They are \emph{public domain}, \emph{Creative Commons}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License}, \emph{GNU Lesser General Public License}, \emph{GNU Free Documentation License}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
\end{itemize}

@@

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
%The schema described above was adopted to create RDF triple structures.
%is a good starting point most of the relations were taken over introducing a PanLex RDF vocabulary.
The entities and relations of the schema described in the previous section
serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside
in the namespace \texttt{<http://panlex.org/plx/>}, abbreviated with \texttt{plx}.
An example of the resulting ontology is depicted in \autoref{tbl:vocabulary} and
summarized as follows:
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern
\emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case
and the \{id\} is the primary key of the corresponding database table.

%As in the original PanLex database there are also \emph{expressions} used in the RDF dataset.
Expressions are modeled as intances of \vocab{plx:Expression}.
Their textual representations and normalized versions thereof become the 
values of the properties \texttt{rdfs:label} and \texttt{plx:degradedText}, respectively.
Their corresponding language variety is stated
using \texttt{plx:languageVariety}.

%These resources have an \texttt{rdfs:label} containing the textual representation of the considered expression.
%Additionally there is a normalized version of the label, reachable via the \texttt{plx:degradedText} property.
%Expression entities also have an assigned language variety (\texttt{plx:languageVariety}).
%It refers to the language variety the textual representation is written in.

For Language and Language Varieties the classes
texttt{plx:Language} and \texttt{plx:LanguageVariety} are introduced.

\emph{ISO 639-1} and \emph{ISO 639-3 codes} are of the classes
\texttt{plx:Iso639-1Code} and \texttt{plx:Iso639-3Code} respectively.

@@

All approvers share the \texttt{plx:Approver} class. The characteristics of an approver are described using mainly triples with literal objects.
These are for example \texttt{plx:registrationDate} pointing to an \texttt{xsd:date} literal,
\texttt{dc:title} to assign the title of a source,
\texttt{dc:creator} to give an \texttt{xsd:string} containing the author's name.

%These entities are mainly used for interlinking to the Lexvo dataset\footnote{\url{http://lexvo.org/ontology}}.
%See \autoref{sec:linking} for more details.

%Besides the approvers there are also \emph{languages} and \emph{language varieties} taken over into the RDF vocabulary.
%They are of the classes \texttt{plx:Language} and \texttt{plx:LanguageVariety} respectively.
%Language varieties have a name given by the \texttt{rdfs:label} property and belong to a certain language referred to via \texttt{plx:languageVarietyOf}.
%For those languages that have an ISO 639-3 language code, this code is referenced via the \texttt{plx:iso639-3Code} property.
%The equivalent ISO 639-1 code can be retrieved using the \texttt{plx:iso639-1Code} property.

%The structure of \emph{approvers} is nearly the same as in the database schema.
% etc.
%All properties defined for \texttt{plx:Approver} instances are shown in \autoref{tbl:vocabulary}.

%Besides the approvers there are also \emph{languages} and \emph{language varieties} taken over into the RDF vocabulary.
%They are of the classes \texttt{plx:Language} and \texttt{plx:LanguageVariety} respectively.
%Language varieties have a name given by the \texttt{rdfs:label} property and belong to a certain language referred to via \texttt{plx:languageVarietyOf}.
%For those languages that have an ISO 639-3 language code, this code is referenced via the \texttt{plx:iso639-3Code} property.
%The equivalent ISO 639-1 code can be retrieved using the \texttt{plx:iso639-1Code} property.

Since the PanLex project compiled its database by extracting data from different sources, the licenses of these sources were also considered.
At present, we support the different license categories given in the database by creating resources of the \texttt{plx:License} class.
%with appropriate descriptions in their \texttt{rdfs:label}.
%In the RDF dataset these have an additional \texttt{rdfs:label} property giving the names or descriptions 
\begin{table}
\begin{scriptsize}
\begin{tabular}{ll}
Class & Properties \\
\toprule
\texttt{plx:Approver} & \texttt{plx:registrationDate} \\
                      & \texttt{rdfs:label} \\
                      & \texttt{foaf:homepage} \\
                      & \texttt{plx:isbn} \\
                      & \texttt{dc:creator} \\
                      & \texttt{dc:title} \\
                      & \texttt{dc:publisher} \\
                      & \texttt{plx:yearOfPublication} \\
                      & \texttt{plx:quality} \\
                      & \texttt{plx:license} \\
\midrule
\texttt{plx:Language}
                      & \texttt{plx:iso639-3Code} \\
                      & \texttt{plx:iso639-1Code} \\
\midrule
\texttt{plx:LanguageVariety}
                      & \texttt{plx:languageVarietyOf} \\
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:Iso639-1Code} & \\
\midrule
\texttt{plx:Iso639-3Code} & \\
\midrule
\texttt{plx:Expression}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{rdfs:label} \\
                      & \texttt{plx:degradedText} \\
\midrule
\texttt{plx:Meaning}
                      & \texttt{plx:approver} \\
                      & \texttt{plx:meaningDefinition} \\
                      & \texttt{plx:identifier} \\
\midrule
\texttt{plx:Definition}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:Denotation}
                      & \texttt{plx:denotationMeaning} \\
                      & \texttt{plx:denotationExpression} \\
                      & \texttt{plx:wordClass} \\
\midrule
\texttt{plx:WordClass}  
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:License}  & \texttt{rdfs:label} \\
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Classes and properties used in the PanLex RDF vocabulary. Note that all \texttt{rdf:type} and \texttt{rdfs:label} properties have been omitted for brevity.}
\label{tbl:vocabulary}
\end{table}

An overview of the PanLex vocabulary is shown in \autoref{fig:vocabulary}.%
\begin{figure}
\includegraphics[width=0.47\textwidth]{images/pdf/ontology.pdf}
\caption{Overview of the PanLex RDF vocabulary}
\label{fig:vocabulary}
\end{figure}

\subsection{Transformation workflow}
\label{sec:conversion}
\todo{how often in PanLex updated?}
The size of the PanLex database (including indexes) is approximately 18 GB,
which is big enough to make recurrent RDF conversion cumbersome:
A full RDF conversion bloats the dataset and results in data duplication.   
Using conventional hardware, it takes impractically long
to convert all the data. This makes
testing and debugging and fixing issues in the modelling or data conversion of the
data very time-consuming.
As the PanLex data already resides in a relational database, the use of a virtual
RDB 2 RDF mapping solution is a natural choice. The \emph{Sparqlify Mapping Language}
offers a very easy-to-use mapping language.
Essentially, these mappings consist of three clauses:
\begin{itemize}
  \item \emph{From:} Specifies the logical SQL table (i.e. table, view or query) underlying the Sparqlify-ML view.
  \item \emph{With:} Defines SPARQL variables by means of expressions over relational columns that yield RDF terms.
  \item \emph{Construct:} The template (i.e. the set of triple patterns) to be generated in this view, based on the SPARQL variable definitions.
\end{itemize}
~\autoref{fig:ex:sparqlify-ml} shows an example of a Sparqlify-ML view for the
languages in PanLex: From each row of the \emph{i1} table three resources are
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language}in the PanLex and the \emph{schema.org} namespace.
\todo{Maybe we should link to lexvo directly.}
This view-based approach also demonstrates that changing the vocubulary or
adding support for new ones does not require an ETL process, and can therefore be done with little effort.
%PanLex's languages  iso 639-3 code 
%typed with plx:Language
%resources are created for
%the languages in PanLex, linked with \emph{lexvo}, and

@@

\begin{figure}
\centering
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
Create View i1 As
  Construct {
    ?lang a plx:Language ;
          a <http://schema.org/Language> ;
          plx:iso639-3Code ?iso3 .
